```{r}
#install.packages("dplyr")
library(dplyr)
library(ggplot2)
```

we will be practicing with `dplyr` functions using the `starwars` dataset:

```{r}
starwars
head(starwars)
?starwars
```

## subset rows with filter

`filter()` allows us to subset the **rows** of a dataset based on the values

```{r}
## fitler for rows with human star wars characters
filter(starwars, species == "Human")

## save to a new R object
humans <- filter(starwars, species == "Human")
filter(humans, sex == "male")
## filter multiple colums at once
filter(starwars, species == "Human", sex =="male")
```

we can use different comparison operators to filter

-   `!` is "not"

-   `==` is "equal"

-   `&` is 'and"

-   `|` is "or"

-   `<` is "less than"

-   `>` is "greater than"

-   i`s.na()` returns true is value is NA

    ```{r}
    ## return characters with a missing mass
    filter(starwars, is.na(mass))
    ```

```{r}
## How to get all characters that are not human
filter(starwars, species != "Human")
```

variable_name`%in%` vector_of_values returns true if the value of variable_name is equal to one of the values in vector_of_values

This is a shortcut fot variable_name == vector *of*value [1] variable_name == vector_of_variable[2]...

```{r}
skin_color <- "grey"
skin_color  %in% c("yellow","fair", "tan") #returns false because "grey" is not in vector of colors

## How would we filter for charactrs with brown or blond hair
filter(starwars, hair_color %in% c("blond", "brown"))
```

grepl("characters_of_interest", variable_name) returns true if the value of variable_name contains the characters_of_interest

```{r}
filter(starwars, grepl("blond", hair_color))

## How would we get all characters with blue in their eyes
filter(starwars, grepl("blue", eye_color))
```

You can use \<, \>, \<=,\>= to compare numeric variables or categorical variables

-   nominal categorical variables (character vectors) are ranked alphabetically

-   ordinal categorical variable (factors) are ranked based on the built-in-order

```{r}
## starwar chacters with a name starting with letter before "D"
filter(starwars, name < "D") #ranking is baed on alphabetical order

str(diamonds)
filter(diamonds, cut > "Good") #ranking is based on levels of the ordered factor cut

"D" < "d" #false
"d" < "D" #true
```

`between`(x, left, right) is a shorthand for x \> = left & \< = right

```{r}
between(5,1,7) # true because e is less than or equal  7 and greater than or equal to 1
filter(starwars, between(mass, 50, 100))
```

### Practice!

Find all starwars characters that

1.  Are missing a hair color
2.  Are shorter than 100cm and are not droids
3.  Were born between the years 100 and 200

```{r}
#1
 filter(starwars, is.na(hair_color))
#2
 filter(starwars, species != "Droid", height < 100) 
#3 
 filter(starwars, between(birth_year, 100, 200))
```

### Order rows with `arrange`

arranged() reorders rows based on the values of column(s)

```{r}
## order by character names
arrange(starwars, name) 
# order of character values: aAbBcCdD

## order by species and birth_year
arrange(starwars, species, birth_year)

## use desc() to get birth_year in descending order 
arrange(starwars, species, desc(birth_year))
```

Practice!!

1.  Arrange the starwars dataset from tallest to shortest characters
2.  Arrange the starwars dataset from shortest to tallest species
3.  Alphabstize the starwars characters by their hair color then eye color

```{r}
#1
arrange(starwars, desc(height))
#2
arrange(starwars, height, species) #order by height, tiebreaker is alphabetical species
#3
arrange(starwars, hair_color , eye_color)
```

#### Subset columns with `select()`

```{r}
select(starwars, hair_color, skin_color, eye_color)
character_color <- select(starwars, hair_color, skin_color, eye_color)
character_color
```

We can also select ultiple columns in a row

```{r}
# all columns between name aand eye_color (inclusive)
select(starwars, name: eye_color)
```

We can also remove columns by name

```{r}
starwars
select(starwars, -films, -vehicles, - starships)
select(starwars, -c(films, vehicles, starships))
select(starwars, -c(films:starships))
```

Helpful Functions

`everything()` is a shortcut for "everything else"

```{r}
select(starwars, species, everything()) # put species first, then all the other columns

## alphabetize columns
names(starwars)
starwars_names <- sort(names(starwars))
starwars_names
select(starwars, starwars_names)
```

`starts_with("abc")` matches columns that start with "abc"

`ends_with("abc")` matches columns that end with "abc"

`contains("abc")` matches columns that contain "abc"

```{r}
select(starwars, starts_with("h"))
select(starwars, contains("or"))

```

Practice!!!

1.  Select each character's homeworld and species. What question would this subset of variables help answer?
2.  Brainstorm as many ways as possible to select hair_color, skin_color, eye_color, and birth_year.

```{r}
#1
select(starwars, homeworld, species)
#2
select(starwars, contains("_"))
select(starwars, contains("color"), birth_year)
select(starwars, hair_color: birth_year)
select(starwars, hair_color, skin_color, eye_color, birth_year)
select(starwars)
```

#1 : What planets have the highest population of a certain species? , what type of species live in the select homeworld? , Where do species live?, How many X species live in X homeworld?

#### Add new variables with `mutate`

```{r}
starwars_small <- select(starwars, height, mass, species)
starwars_small

mutate(starwars_small,
       height_m = height/100)
```

We can also create multiple variables at once and we can refer to variable we just created

```{r}
mutate(starwars_small,
       height_m = height/100,
       bmi = mass/height_m^2)

## only keep new variables
transmute(starwars_small,
          height_m = height/100,
          bmi = mass/height_m^2)
```

Summary statistics give one value for the entire column: sum(), mean()

```{r}
mutate(starwars_small,
       sum = sum(mass, na.rm = T))

mutate(starwars_small, 
       row_addition = height + mass ,
       sumary_sum = sum(height , mass, na.rm = T))
```

Group by variable using the `.by` parameter

```{r}


```

Change the type of a variable

```{r}
mutate(starwars_small,
       mass= as.character(mass))
```

#### Summarize data with `summarize`

```{r}
summarize(starwars,
          mean_mass = mean(mass, na.rm = T))

#Group by species
summarize(starwars,
          mean_mass = mean(mass, na.rm = T),
          .by = species)
```

The only columns kept are the columns you group by and the columns you create.

n() is a helpful function thay returns the number of rows in a group

```{r}
summarize(starwars,
          mean_mass = mean(mass, na.rm = T),
          count = n(),
          .by = species)

##group by multiple groups
summarize(starwars,
          mean_mass = mean(mass, na.rm = T),
          count = n(),
          .by = c(species, sex))
```
